Chemical tagging can work: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity

### 2016-08-06 papers/kmeans/response_to_referee.txt

Dear Editor:

Please find below the response to our (very helpful) referee report on
manuscript ApJ/ApJ101852.  We have taken all of the referee's
suggestions, and made some other changes from comments we have
received from colleagues.

@@

    There are some missing references to the original demonstrations
    of chemical tagging of unbound groups, namely the Wolf 630 group
    and HR 1614 group. Those were key to building up the case for
    this "tagging" research.

We have added the following references: De Silva et al (2007), Eggen
(1970), and Bubar & King (2010).

    The text needs to highlight that the training set is critical -
    Cannon doesn't know physics. It needs to be told what is "true"
    and that the subsequent results will be limited to the believed
    level of truth. While it should be specified that tagging per se
    doesn't require absolute abundances, but the subsequent
    interpretation of the chemical accretion history will require
    "true" abundances - not just precision. Also some clarification
    about the Cannon's performance at the boundaries of the training
    set parameters would have been good to see.

This statement and point is true and wise. We have modified the text
in [HOGG]

@@

We have added references to: Norris & Da Costa (1995), which (to our
knowledge) was the first to unambiguously show anti- correlations in
Na, Al, and O, and to Carretta et al. (2009), which provides a more
recent and comprehensive view of our current knowledge of elemental
abundance correlations in gloublar clusters.

    Section 3:

    The text doesn't clarify why only k-means was chosen to do the
    'tagging' given the issues discussed, was it simply easy and
    fast? Fine as the focus of the paper is the precision of the
    abundance results with Cannon, not the fidelity of group finding.

We have added text to clarify why k-means was chosen. It is
fundamentally because k-means is the clustering algorithm of "first
resort", which contains very few assumptions and only one (integer)
adjustable parameter (not counting the metric on the vector space).

@@

HOGG

-----
Other changes:

In addition to all the above, we made the following additional
changes:

Because of discussions and checking of results that happened
subsequent to our submission of this paper with the APOGEE team, some
of which made use of proprietary data and information about the
survey, we have expanded the author list in accordance with the
APOGEE/SDSS-III authorship policies.  The consequence is that there
are now several additional new authors, most of whom contributed to
this project most directly by building the APOGEE project (and hence
the input data and tests thereof).

We added the two crucial APOGEE references: Eisenstein 2011 (SDSS-III)
Holtzman 2015 (APOGEE DR12).

@@

HOGG:  Adjusted what we say about Ting et al?

HOGG:  Commented on, eg, the Na abundances?

HOGG:  Go through github issues and emails?

### 2016-08-06 papers/kmeans/kmeans.tex
omg figurename too?
frenchspacing
nothing
authors, affils, response text

% omg AASTEX 6.0
\setlength{\voffset}{0in}
\setlength{\hoffset}{0in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{9in}
\setlength{\headheight}{0ex}
\setlength{\headsep}{\baselineskip} % this is 2 lines in ``manuscript''
\setlength{\topmargin}{-1\headsep}
\linespread{0.54} % close to 10/13 spacing in ``manuscript''
\setlength{\parindent}{0.54\baselineskip}
\hypersetup{
  colorlinks = false,
}
\makeatletter % you know you are living your life wrong when you need to do this
\long\def\frontmatter@title@above{
\vspace*{-\headsep}\vspace*{\headheight}
\noindent\footnotesize
{\noindent\footnotesize\textsc{\@journalinfo}}\par
{\noindent\scriptsize Preprint typeset using \LaTeX\ style AASTeX6 v.\ 1.0 but heavily modified by David W. Hogg (NYU).
}\par\vspace*{-\baselineskip}\vspace*{0.625in}
}%
\makeatother

% text
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\sdssiii}{\project{\acronym{SDSS-III}}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\aspcap}{\project{\acronym{ASPCAP}}}
\newcommand{\gaiaeso}{\project{Gaia--\acronym{ESO}}}
\newcommand{\galah}{\project{\acronym{GALAH}}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\etal}{\foreign{et~al.}}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}

@@

\title{Chemical tagging can work: \\
       Identification of stellar phase-space structures \\
       purely by chemical-abundance similarity}
\author{David~W.~Hogg\altaffilmark{1,2,3,4},
        Andrew~R.~Casey\altaffilmark{5},
        Melissa~Ness\altaffilmark{4},
        Hans-Walter~Rix\altaffilmark{4},
        Daniel~Foreman-Mackey\altaffilmark{6,7},
        Anna~Y.~Q.~Ho\altaffilmark{8},
        Jon~A.~Holtzman\altaffilmark{9},
        David~L.~Nidever\altaffilmark{10}}
\altaffiltext{1}{Simons Center for Data Analysis, 160 Fifth Avenue, 7th floor, New York, NY 10010, USA}
\altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Physics,
  New York University, 4 Washington Pl., room 424, New York, NY 10003, USA}
\altaffiltext{3}{Center for Data Science, New York University, 726 Broadway, 7th floor, New York, NY 10003, USA}
\altaffiltext{4}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
\altaffiltext{5}{Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3~0HA, UK}
\altaffiltext{6}{Sagan Fellow}
\altaffiltext{7}{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195, USA}
\altaffiltext{8}{Astronomy Department, California Institute of Technology, MC 249-17, 1200 East California Blvd, Pasadena, CA 91125, USA}
\altaffiltext{9}{New Mexico State University, Las Cruces, NM 88003, USA}
\altaffiltext{10}{Steward Observatory, 933 North Cherry Ave, Tuscon, AZ 85719, USA}

\begin{abstract}
% Context:
Chemical tagging promises to use detailed abundance measurements to
identify spatially separated stars that were in fact born together
(in the same molecular cloud), long ago.
This idea has not previously yielded scientific successes, probably because of the
noise and incompleteness in chemical-abundance measurements.
However, we have succeeded in substantially improving spectroscopic
measurements with \thecannon, which has delivered $15$ individual
abundances for $\sim 10^5$ stars observed as part of the
\apogee\ spectroscopic survey, with precisions around 0.04~dex.
% Aims:
We test the chemical-tagging hypothesis by looking at clusters in abundance space
and confirming that they are clustered in phase space.
% Methods:
We identify (by the k-means algorithm) overdensities of stars
in the 15-dimensional chemical-abundance space delivered by \thecannon,
and plot the associated stars in phase space.
We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
% Results:
We find that clusters in abundance space are indeed clusters in phase space.
We recover some known phase-space clusters and find other interesting structures.
This confirms the chemical-tagging hypothesis and verifies the precision of the abundance measurements delivered by \thecannon.
This is the first-ever project to identify phase-space structures at survey-scale by blind search purely in abundance space;
the prospects for future data sets are very good.
\end{abstract}

@@

The abundance-space clusters shown in the \figurename s
might not look dense in 2-d projections of the abundance space, but they
are very dense in the 15-dimensional space.  Indeed,
\figurename~\ref{fig:M13} is the densest cluster found in the 15-d
space \emph{by far}.
The challenge of this work is to find structure in the high
dimensionality space that is not obviously visible in any 2-d projection.
Although we have by no means any model of the 15-d space, we do know
that the structures shown in the figures are very high in 15-d
density.

@@

\acknowledgements
It is a pleasure to thank
  Jeremy~Magland (\acronym{SCDA}),
  Don Schneider (PSU), and
  the Blanton--Hogg group meeting
for valuable comments and discussions.
This project was funded in part by
  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
  the European Research Council under the
  European Union's Seventh Framework Programme (FP~7)
  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
This research made use of the \acronym{NASA} \project{Astrophysics Data System}.

>>>>

% omg AASTEX 6.0
\setlength{\voffset}{0in}
\setlength{\hoffset}{0in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{9in}
\setlength{\headheight}{0ex}
\setlength{\headsep}{\baselineskip} % this is 2 lines in ``manuscript''
\setlength{\topmargin}{-1\headsep}
\linespread{0.54} % close to 10/13 spacing in ``manuscript''
\setlength{\parindent}{0.54\baselineskip}
\hypersetup{
  colorlinks = false,
}
\makeatletter % you know you are living your life wrong when you need to do this
\long\def\frontmatter@title@above{
\vspace*{-\headsep}\vspace*{\headheight}
\noindent\footnotesize
{\noindent\footnotesize\textsc{\@journalinfo}}\par
{\noindent\scriptsize Preprint typeset using \LaTeX\ style AASTeX6 v.\ 1.0 but heavily modified by David W. Hogg (NYU).
}\par\vspace*{-\baselineskip}\vspace*{0.625in}
}%
\makeatother
\frenchspacing
\sloppy\sloppypar

% text
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\sdssiii}{\project{\acronym{SDSS-III}}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\aspcap}{\project{\acronym{ASPCAP}}}
\newcommand{\gaiaeso}{\project{Gaia--\acronym{ESO}}}
\newcommand{\galah}{\project{\acronym{GALAH}}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\etal}{\foreign{et~al.}}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\renewcommand{\figurename}{Figure} % omg AASTEX 6

@@

\title{Chemical tagging can work: \\
       Identification of stellar phase-space structures \\
       purely by chemical-abundance similarity}
\author{David~W.~Hogg\altaffilmark{1,2,3,4},
        Andrew~R.~Casey\altaffilmark{5},
        Melissa~Ness\altaffilmark{4},
        Hans-Walter~Rix\altaffilmark{4},
        Daniel~Foreman-Mackey\altaffilmark{6,7},
        Sten~Hasselquist\altaffilmark{8,9},
        Anna~Y.~Q.~Ho\altaffilmark{10},
        Jon~A.~Holtzman\altaffilmark{8},
        Steven~R.~Majewski\altaffilmark{9},
        Sarah~L.~Martell\altaffilmark{11},
        Szabolcs~M\'esz\'aros\altaffilmark{12},
        David~L.~Nidever\altaffilmark{13},
        Matthew~Shetrone\altaffilmark{14}}
\altaffiltext{1}{Simons Center for Data Analysis, 160 Fifth Avenue, 7th floor, New York, NY 10010, USA}
\altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Physics,
  New York University, 4 Washington Pl., room 424, New York, NY 10003, USA}
\altaffiltext{3}{Center for Data Science, New York University, 726 Broadway, 7th floor, New York, NY 10003, USA}
\altaffiltext{4}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
\altaffiltext{5}{Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3~0HA, UK}
\altaffiltext{6}{Sagan Fellow}
\altaffiltext{7}{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195, USA}
\altaffiltext{8}{New Mexico State University, Las Cruces, NM 88003, USA}
\altaffiltext{9}{Department of Astronomy, University of Virginia, Charlottesville, VA 22904-4325, USA}
\altaffiltext{10}{Astronomy Department, California Institute of Technology, MC 249-17, 1200 East California Blvd, Pasadena, CA 91125, USA}
\altaffiltext{11}{School of Physics, University of South Wales, Sydney 2052, Australia}
\altaffiltext{12}{ELTE Gothard Astrophysical Observatory, H-9704 Szombathely, Szent Imre Herceg st. 112, Hungary}
\altaffiltext{13}{Steward Observatory, 933 North Cherry Ave, Tuscon, AZ 85719, USA}
\altaffiltext{14}{University of Texas at Austin, McDonald Observatory, USA}

\begin{abstract}
% Context:
Chemical tagging promises to use detailed abundance measurements to
identify spatially separated stars that were in fact born together
(in the same molecular cloud), long ago.
This idea has not previously yielded scientific successes, probably because of the
noise and incompleteness in chemical-abundance measurements.
However, we have succeeded in substantially improving spectroscopic
measurements with \thecannon, which has now delivered $15$ individual
abundances for $\sim 10^5$ stars observed as part of the
\apogee\ spectroscopic survey, with precisions around 0.04~dex.
% Aims:
We test the chemical-tagging hypothesis by looking at clusters in abundance space
and confirming that they are clustered in phase space.
% Methods:
We identify (by the k-means algorithm) overdensities of stars
in the 15-dimensional chemical-abundance space delivered by \thecannon,
and plot the associated stars in phase space.
We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
% Results:
We find that clusters in abundance space are indeed clusters in phase space.
We recover some known phase-space clusters and find other interesting structures.
This confirms the chemical-tagging hypothesis and verifies the precision of the abundance measurements delivered by \thecannon.
This is the first-ever project to identify phase-space structures at survey-scale by blind search purely in abundance space;
the prospects for future data sets are very good.
\end{abstract}

@@

The abundance-space clusters shown in the \figurename s
might not look dense in 2-d projections of the abundance space, but they
are very dense in the 15-dimensional space.  Indeed,
\figurename~\ref{fig:M13} is the densest cluster found in the 15-d
space \emph{by far}.
The challenge of this work is to find structure in the high
dimensionality space that is not obviously visible in any 2-d projection.
Although we have by no means any model of the 15-d space, we do know
that the structures shown in the \figurename s are very high in 15-d
density.

@@

\acknowledgements
It is a pleasure to thank
  Katia Cunha (NOAO),
  Jeremy Magland (\acronym{SCDA}),
  Don Schneider (PSU),
  Keivan Stassun (Vanderbilt),
  and the Blanton--Hogg group meeting
for valuable comments and discussions.
This project was funded in part by
  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1312863}, \acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
  the European Research Council under the
  European Union's Seventh Framework Programme (FP~7)
  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
This research made use of the \acronym{NASA} \project{Astrophysics Data System}.
